---
title: "Seasonality and spatial context shape the gut microbiome diversity of the common carder bee *Bombus pascuorum*"
author: "Luis Javier Chueca^1^*, Jon Poza^1^, Manpreet K. Dhami^2,3^, Marion Donald^2^, Christian Gostout^1^, Brais Hermosilla^1^, Jennifer Rose^1^, Xabier Salgado-Irazabal^1^, Ainhoa Magrach^1,4^"
corresponding-author: "Luis Javier Chueca (luisjavier.chueca@bc3research.org)"
format: docx
editor: visual
bibliography: references.bib
csl: molecular-ecology.csl
---

\small

```         
^1^Basque Centre for Climate Change-BC3, Edif. Sede 1, 1°, Parque Científico UPV-EHU, Barrio Sarriena s/n, 48940 Leioa, Spain

^2^Manaaki Whenua Landcare Research, Bioeconomy Science Institute, Lincoln, New Zealand\

^3^School of Biological Sciences, Waipapa Taumata Rau, University of Auckland, Auckland, New Zealand\

^4^IKERBASQUE, Basque Foundation for Science, María Díaz de Haro 3,48013 Bilbao, Spain
```

## Abstract

Pollinators are essential for both natural and agricultural systems, supporting the diversity of plant communities and the productivity of crops around the globe. However, different global change drivers increasingly threaten these pollinator populations, leading to declines in population numbers and cascading effects on ecosystems and food security. One critical but understudied aspect of pollinator health is their microbiome. The microbiome, a community of microorganisms living within and on pollinators, plays a significant role in digestion, immune function, and disease resistance, amongst others. While different environmental changes can disrupt these microbial communities, compromising pollinator health and survival, most research on pollinator microbiomes has focused on managed species like honeybees and short temporal periods, leaving significant gaps in our understanding of wild pollinator microbiomes over longer time periods. This study investigates the microbiome composition of the common carder bee (*Bombus pascuorum*) across different floral environments throughout the flowering season. We show that gut microbiomes are highly dynamic yet structured, with bacterial diversity primarily shaped by seasonal dynamics and overall plant diversity, whereas fungal diversity is strongly site-filtered and closely associated with diet composition. These results reveal distinct spatiotemporal drivers of bacterial and fungal communities and highlight the importance of considering both temporal and environmental context when assessing pollinator microbiomes. Understanding how wild pollinator microbiomes respond to environmental variability is essential for predicting their resilience to global change and informing conservation strategies.

## Introduction

Pollinators play a critical role in the functioning of both natural and agricultural systems worldwide, facilitating the reproduction of \>85% of flowering plants [@ollerton2011] and contributing to the production of \>75% of global food crops [@klein2006]. However, pollinators are increasingly at risk due to multiple global change drivers, something that has been recently defined as "*death by a thousand cuts*" [@wagner2021]. Climate change, with its impacts on temperature and precipitation patterns, disrupts pollinator behaviors and/or morphologies [@miller-struttmann2015] leading to plant-pollinator mismatches [@gérard2020]. Habitat loss and fragmentation lead to a reduction in pollinator populations [@winfree2009], while the widespread use of pesticides introduce additional stresses [@guzman2024]. These factors, compounded by the spread of diseases [@manley2015] and the increasing presence of invasive species [@vanbergen2017], threaten the survival of pollinator populations, leading to cascading effects on ecosystems and human food systems [@devkota2024; @murphy2022].

A relatively unexplored yet crucial aspect of pollinator populations is their microbiome [@shah2024], the community of microorganisms living within and on them. The microbiome can significantly impact the health, physiology [@Engel2013], development, and survival [@ferguson2018] of organisms, influencing digestion, immune function, and disease resistance. In insect pollinators, the gut microbiome plays a vital role in nutrient absorption [@zheng2017; @zheng2019], detoxification of harmful substances [@engel2012], and protection against pathogens [@steele2021; @koch2011]. Disruptions to the microbiome can compromise these essential functions and lead to increased vulnerability and mortality [@maes2016]. Understanding the composition and functionality of microbiomes is therefore critical, not only for gaining insights into the biology and ecology of pollinator species but also for developing strategies to mitigate the adverse effects of global change drivers.

A lot of the research on pollinator microbiomes has focused on managed species, like honeybees [@engel2012] and certain bumblebee species [@weinhold2024], providing valuable insights into their gut microbial communities and how these microorganisms influence their health and productivity. However, our understanding of the microbiomes of other pollinator species, and particularly of wild species, remains limited (although see [@li2023; @mayr2021; @mcfrederick2016; @voulgari-kokota2019] for a focus on other species). This bias hampers our ability to generalize findings beyond model taxa, limits the integration of microbiomes into conservation and management strategies for wild pollinators, and constrains our understanding of their potential roles in pollination processes and ecosystem stability. We know even less about how these microbiomes change over time or how different environmental factors shape them. As global change continues to alter ecosystems, understanding the temporal and spatial dynamics of pollinator microbiomes could provide valuable insights into how these species might cope with environmental stresses, maintain ecosystem functions, and contribute to biodiversity conservation.

Research in honeybees and some bumblebees has shown the existence of a core microbiome in these species, defined as a common group of microbes that are found across individuals of a given species (although the existence of this core microbiome is still somewhat controversial [@risely2020; @neu2021]. This core microbiome consists of around five genera in the case of honeybees [@carlini2024] and is mostly vertically-transmitted within the hive [@martinson2010]. In bumblebees, evidence also points to the presence of a relatively stable core microbiome, typically dominated by Snodgrassella, Gilliamella, Bifidobacterium, and Lactobacillus spp. [@kwong2017; @li2021].

However, pollinator microbiomes are shaped by multiple pathways of microbial acquisition, including vertical or social transmission within colonies as well as horizontal acquisition from the environment. In bumblebees, colonies are annual and founded each season by a single queen, which may increase reliance on both social interactions and environmentally acquired microbes for establishing and maintaining the gut microbiome. Beyond the core microbiome shared among individuals, pollinators can acquire additional microbes through their diet, particularly via nectar and pollen from floral resources [@ludvigsen2015]. Because floral resource availability and composition change over the flowering season and differ among sites, these temporal and spatial dynamics may strongly influence both the plant material ingested by bees and the diversity and composition of their gut microbial communities. These acquired microorganisms can provide pollinators with new functions and adaptations [@martin2022] and may also reduce their susceptibility to environmental changes and stressors [@shah2024]. Moreover, variation in floral resource diversity and diet breadth can further influence microbiome structure, as bees foraging on a wider range of plant species encounter a greater diversity of microbes and metabolites that may persist in the gut [@bosmans2018; @ptaszynska2021; @lignon2024]. This highlights the importance of understanding how different environmental variables shape microbiome composition—for example, climatic conditions [@castelli2021; @papp2022], seasonality [@kesnerová2019; @almeida2022; @li2022], individual age [@maes2021], habitat type [@jones2017; @westreich2022], diet [@geldert2020], and/or floral resource availability [@meehan2025].

Here, we focus on understanding the gut microbiome composition and dynamics of a wild pollinator species, the common carder bee (*Bombus pascuorum*), throughout the flowering season as well as across different sites that vary in floral diversity. By jointly analyzing gut microbial communities and plant taxa detected in the gut across contrasting floral environments and throughout the flowering season, we aim to determine how seasonal dynamics, environmental plant diversity, and realized diet shape the structure and diversity of bacterial and fungal gut microbiomes in a wild bumblebee. This study sheds light on the temporal and spatial changes in gut microbiome composition within *B. pascuorum* and reveals how different ecological contexts shape these vital microbial communities. Our findings contribute to a deeper understanding of the ecological interactions between wild pollinators and their environments, providing insights that are essential for conservation efforts and the maintenance of healthy pollinator populations.

## Methods

Our study was conducted within Gorbeia Natural Park, a protected area located in the Basque Country in N Spain. The region is shaped by the Atlantic Ocean's influence, resulting in mild winters and summers, alongside high annual rainfall, averaging around 1300 mm. During the study period (April to July), the average temperatures range from 10°C to 17.5°C [@Pizarro2021]. The area's vegetation includes beech forests mixed with extensive conifer plantations. Additionally, there is a notable presence of shrubland and meadows, particularly in the rockier, steeper regions of the mountain [@Albizu2002COMPARACINDL].

Within the park, we selected 16 different sampling sites, all of them located at similar altitudes \~800 m a.s.l, which we sampled every two weeks depending on weather conditions, avoiding rainy, windy or foggy days. At every sampling date (from now on referred to as sampling period), we aimed to collect at least 5 *B. pascuorum* individuals, although this was not always possible.

*B. pascuorum* is a relatively generalist forager (Artamendi et al in prep) with colony sizes that are moderate in size compared to other bumblebee species, ranging from 50-200 individuals [@herrmann2007]. For our sampling we specifically focused on foraging worker females—the individuals actively collecting nectar and pollen during the field season. All collected individuals were labeled and stored independently in 50 ml vials at -20ºC until processed in the laboratory. We then extracted bumblebee guts using sterile forceps. Guts were stored in 96% ethanol until DNA extraction.
In addition, within each site we conducted a number of floral surveys each time we sampled the area following a 250 m long linear transect where we counted the number of flowers for all flowering species.

*DNA Extraction and PCR Amplification*

We extracted bacterial, plant, and fungal DNA using the NucleoSpin® 96 Soil kit (Macherey-Nagel, Düren, Germany) and amplified them in duplicate using the DFD forward and ASDFAS reverse primers. To avoid any site or period bias, we randomly organized all samples on extraction plates before DNA extraction, creating a new individual code for each sample. Each gut sample was placed in a separate 2 mL vial with ceramic beads, and we allowed the ethanol to evaporate completely. Once dry, we added 700 µL of Lysis Buffer (SL1) to each vial and mechanically homogenized the samples for 10 minutes using a Genogrinder. Following homogenization, we centrifuged the samples at 11,000 × g for 2 minutes to remove foam. We then added 150 µL of Buffer SL3 to each sample, vortexed for 5 seconds, incubated on ice (0–4 ºC) for 5 minutes, and centrifuged again at 11,000 × g for 1 minute. We filtered the lysate using a square-well block and the NucleoSpin® Inhibitor Removal Plate. We loaded 550 µL of the clear supernatant into each well and centrifuged at 2,250 × g for 13 minutes. After discarding the NucleoSpin® Inhibitor Removal Plate, we added 250 µL of Buffer SB to the flow-through and mixed it by pipetting. To bind the DNA, we transferred 725 µL of each sample to a NucleoSpin® Soil Binding Plate and centrifuged it at 2,250 × g for 13 minutes, discarding the flow-through afterward.We washed the DNA in four sequential steps: (1) 500 µL of Buffer SB, centrifuged at 2,250 × g for 13 minutes; (2) 550 µL of Buffer SW1, centrifuged at 2,250 × g for 5 minutes; (3) 700 µL of Buffer SW2, centrifuged at 2,250 × g for 5 minutes; and (4) 700 µL of Buffer SW2, centrifuged at 2,250 × g for 13 minutes. After washing, we incubated the plate at 37 ºC for 20 minutes to dry the silica membrane. Finally, we eluted the DNA by adding 100 µL of Buffer SE to each well, incubating for 1 minute, and centrifuging at 2,250 × g for 5 minutes. In total, we obtained 126 samples. To confirm successful DNA extraction, we ran Nanodrop tests on a set of random samples.

*Library preparation and bioinformatic processing*

We characterized the common carder bee’s gut microbiome (bacteria and fungi) and its plant-derived gut contents using community metabarcoding with a two-step PCR followed by high-throughput sequencing. Step 1 PCRs were performed separately for bacterial, plant, and fungal genetic material, with identical reagent volumes for all samples, differing only in the primers used (Table S01). PCRs were run in a 96-well plate, with two plates for each group: one containing 96 samples and another with 40 samples, including negative controls. Step 1 PCR was repeated twice per sample, after which 7.5 µL from each replicate were pooled into a new plate, with 15 µL per well. We cleaned and equalized the pooled PCR products using Omega Mag-Bind Total Pure NGS magnetic beads. Briefly, beads were brought to room temperature and vortexed before adding beads to each well at a 1:1 bead-to-sample ratio. The mixture was pipetted to mix and incubated at room temperature for 5 minutes. Samples were then placed on a magnetic plate for approximately 5 minutes to allow the beads to separate from the solution. We removed the supernatant, and added 180 µL of freshly prepared 85% ethanol for a wash step. After a 30-second incubation, we removed the ethanol, and repeated the wash step. We sealed the plate, centrifuged it briefly in a benchtop plate centrifuge, placed it back on the magnetic plate and pipetted off any remaining ethanol. We allowed the beads to air dry before eluting the cleaned PCR product in 15 µL of 1x TE buffer. We used 1.8 µL of this purified Step 1 PCR product as the template for Step 2 PCRs (add the PCR protocol below or in the supplement) and repeated this bead cleaning and equalization protocol for the Step 2 PCR products. We pooled samples with equal volume to create pooled bacterial, fungal, and plant libraries and assessed the fragment distribution of the pooled libraries via high-sensitivity DNA assay on a bioanalyzer. Pooled libraries were quantified using qubit high sensitivity ds DNA kit (Qubit details), diluted to 4nM and run on the Illumina MiSeq 2100 instrument at the Ecological Genetics Laboratory, Manaaki Whenua Landcare Research, Auckland, using the MiSeq Reagent Kit v3, for 600 cycles in paired-end mode. To obtain Amplicon Sequence Variants (ASVs), we used barcode free, paired-end reads of demultiplexed samples. Illumina reads were quality-filtered, and adapters and low-quality bases were trimmed prior downstream analysis. We processed the reads using the DADA2 pipeline [@callahan2016] in R, which included error model learning, dereplication, denoising, merging of paired-end reads, and chimera removal. The resulting ASVs were taxonomically assigned against a reference databased specific to each target marker: the SILVA_SSU_r138_1 16S rRNA database for bacteria [@quast2012; @yilmaz2013], the UNITE v10 database for fungal ITS1 sequences [@abarenkov2025], and Bell ITS2 database [@bell2021] for plant sequences. To identify and remove potential contaminants, we ran the Decontam package [@davis2018] independently to each ASV dataset, using prevalence-based methods. All settings and parameters used for each database are provided at Supplementary Table S02 and in the project’s GitHub repository.

*Statistical analyses*

To evaluate the completeness of our surveys, we used the iNEXT package [@iNEXT], which provides robust estimates of species diversity and sampling coverage based on sample-size- and coverage-based rarefaction and extrapolation. We compared the observed sample coverage with the estimated asymptotic coverage to quantify the fraction of the community captured by our sampling efforts. We performed these analyses for the bacteria, fungi and plant communities in the gut.

We then visualised spatial and temporal differences in both bacterial and fungal community compositions using Non-metric Multidimensional Scaling (NMDS) with the Bray-Curtis dissimilarity index, implemented in the vegan package in R [@vegan]. ASV abundance data were square-root transformed to reduce the influence of dominant species. We then performed a Permutational Multivariate Analysis of Variance (PERMANOVA) with 999 permutations to identify significant differences across the sampling periods and sites. To evaluate whether dispersion effects could influence the results, we tested for homogeneity of group dispersions using the function *betadisper* [@vegan].In addition, we conducted pairwise PERMANOVA comparisons between sampling periods using repeated *adonis2* tests in the *vegan* package [@vegan], with permutations constrained within site when possible. P-values were adjusted for multiple testing using Benjamini-Hochberg false discovery rate procedure [@benjamini1995].

We quantified β-diversity in bacterial and fungal communities across space and time using the betapart R package [@betapart]. Using presence–absence data, we first summarized overall heterogeneity across all samples by partitioning multi-site β-diversity (β_multi) into turnover (β_JTU; species replacement) and nestedness (β_JNE; richness differences without replacement). We then examined pairwise dissimilarities among samples with beta.pair and related these to spatial and temporal separation. Spatial separation was measured as great-circle distance between sampling coordinates, and temporal separation as the absolute difference in sampling dates (days). Further, we evaluated distance–decay by testing the correlation between pairwise β_JTU and spatial or temporal distances using Mantel tests, and visualized relationships with linear models of pairwise dissimilarities against distance (noting the non-independence of pairwise data). To incorporate abundance information, we repeated these analyses using Bray–Curtis dissimilarities, partitioned with beta.pair.abund into balanced variation in abundances and abundance gradients.

To evaluate how the Shannon diversity of bacteria, fungi and plant species found in the gut of wild common carder bees varied across space and time in relation to floral diversity at the site level, we fitted linear mixed‐effects models with a Gaussian error distribution. Floral species richness was decomposed into a within‐site component (temporal deviations from each site’s mean) and a between‐site component (site mean), allowing us to distinguish temporal from spatial effects. Sampling period was included as a fixed effect to capture additional temporal trends. We verified that predictors were not strongly collinear (all VIFs \< 3). Sequencing depth (log‐transformed read counts) was included as a covariate. To account for repeated sampling and site‐specific temporal dynamics, models included random intercepts and random slopes for period by site. Because observations were taken repeatedly over time, we also tested robustness to temporal autocorrelation using an AR(1) correlation structure; results were consistent across specifications. We also assessed whether gut plant diversity was associated with microbial diversity fitting mixed-effects models (GLMMs, Gaussian) of bacterial or fungal ASV richness as a function of gut plant richness and sampling period, with site as a random effect.

*Co-occurrence network inference*

We inferred co-occurrence patterns to see which taxa covary across bee individuals and study periods—revealing potential interactions, community structure, and different potential plant–microbiome or bacteria-fungi links. After removing low-depth samples (≥300 reads for 16S/ITS; ≥100 for plant metabarcoding where applicable), we retained the 100 most abundant amplicon sequence variants (ASVs) in each dataset to reduce sparsity and facilitate cross-sample comparisons. Count tables were centered log-ratio (CLR) transformed using a pseudocount of 1 to mitigate compositional effects, and pairwise Pearson correlations were computed. To retain potentially ecologically meaningful associations while excluding spurious near-zero values, we kept edges with ∣r∣ \> 0.1 and recorded their sign as positive or negative. Although this is a relatively low threshold, microbial associations are often weak but consistent, and stricter cutoffs would risk discarding relevant ecological structure. Cross-domain networks (bacteria–fungi, plants–bacteria, plants–fungi) were computed on the intersection of samples passing the respective filters, concatenating CLR-transformed matrices before correlation (Bacteria–Fungi: n=91; Plants–Bacteria: n=120; Plants–Fungi: n=88). We defined hubs as nodes with the highest degree (number of edges connected to a node) and weighted strength (sum of the absolute values of correlation coefficients for those edges) in the resulting networks. Correlations are interpreted as hypotheses of association rather than direct interactions.

*Functional Prediction of Microbial Communities*

To infer the functional potential of the bacterial communities, we used PICRUSt2 (Phylogenetic Investigation of Communities by Reconstruction of Unobserved States, version X.X) to predict metagenomic profiles from 16S rRNA gene sequence data. Amplicon sequence variants (ASVs) were aligned to a reference phylogeny using the place_seqsand hsp.py steps, and functional profiles were inferred based on KEGG Orthology (KO) and MetaCyc pathway databases. Predicted pathway abundances were normalized and compared across time periods. Based on observed differences in bacterial diversity and composition over time, we grouped sampling periods into early (March-May) and late (May-July) categories. Statistical comparisons of pathway relative abundances between these two groups were performed using differential abundance analysis, with significance assessed via permutation tests and adjusted p-values. This approach allowed us to assess seasonal shifts in the functional potential of gut microbial communities associated with changes in environmental conditions and floral resources.

## Results

Throughout the flowering season, we collected a total of 126 common carder bee specimens across 16 sites, with an average of 2.58 ± 1.61 individuals per site and sampling period. Sampling completeness was generally high, averaging 90, 98, and 99% across sites and periods in the case of bacteria, fungi and plants respectively, although it ranged widely from 21, 60 and 88% to full completeness (Fig. S1A-C) indicating robust coverage for all community types in our samples.


Examining temporal changes in community composition revealed a stable core microbiome consistently present across all time periods and sampling sites in the case of bacteria (Fig. 1A). This core includes key ASVs from well-known bumblebee-associated genera such as Lactobacillus, Gilliamella, Snodgrassella, and Bombiscardovia, likely reflecting essential symbionts fundamental to host health and function. Interestingly, as the flowering season progressed, this core microbiome was supplemented by additional bacterial taxa, including species from genera such as Pseudomonas, Bacillus, Commensalibacter, and Apibacter, indicating a seasonal enrichment or colonization of the gut by a broader suite of bacteria (Fig. S3). In contrast, fungal communities did not exhibit a persistent core across periods (Fig. 1B). Instead, they were characterized by marked temporal turnover, with different genera dominating at different points in the season. For example, Cladosporium and Alternaria were abundant in some early samples, while Aspergillus, Penicillium, and Fusarium increased in later periods (Fig. S4). This pattern suggests that fungi are more transient members of the bee gut community, likely reflecting opportunistic colonization from the environment rather than stable, host-associated symbionts.

Multivariate analyses of bacterial community composition through time showed clear temporal structuring, with distinct clustering of samples by sampling period. PERMANOVA indicated a significant effect of period (Fig. 2A; F = 3.63, p= 0.0001; R2 = 0.119) and a significant effect of site (F = 1.75, p 0 0.002; R2= 0.173, Fig. S5). Although both factors were significant, site accounted for a slightly larger share of variation than period. Tests for homogeneity of multivariate dispersions indicated no significant differences among periods, confirming that the observed patterns are due to shifts in community centroids rather than differences in within-group variability. Pairwise PERMANOVA comparisons further confirmed that differences were strongest between non-adjacent periods, particularly between the earliest and latest sampling points (Fig. 2B), reinforcing the conclusion that temporal turnover dominates the structuring of gut microbial communities in the common carder bee. Pairwise comparisons revealed that gut microbial communities differed most strongly between mid-season sampling periods and both early and late season communities, whereas early- and late-season microbiomes were comparatively similar. This non-linear temporal pattern suggests that seasonal turnover is primarily driven by transient, low-abundance taxa rather than the core microbiome. Early and late in the season, when floral diversity and dietary breadth are reduced, gut communities appear more strongly structured by a stable core, whereas increased exposure to diverse floral resources during mid-season promotes the acquisition and turnover of peripheral taxa. Multivariate analyses of the fungal dataset showed only subtle temporal structuring and comparatively stronger separation by site. PERMANOVA detected significant effects of both period and site (period: F = 1.30, p = 0.001, R² = 0.053, Fig. S6A; site: F = 1.14, p = 0.002, R² = 0.139, Fig. S7), with site explaining \~2.6× more variation than period (Fig. S7). Residual variation remained high (R² = 0.807), indicating considerable within-group heterogeneity. Overall, these results point to site identity as the dominant driver of the mycobiome composition, with temporal turnover playing a secondary role.

For bacteria, across sites, multi-site Jaccard β-diversity was high (β_JAC = 0.892), largely attributable to turnover (β_JTU = 0.821) rather than nestedness (β_JNE = 0.071), and community dissimilarity showed only a weak, non-significant distance–decay (Mantel r = 0.143, p = 0.085; linear model slope = 0.0125 per km, p = 0.113, R² = 0.021). Across time periods, β-diversity was even higher (β_JAC = 0.985; β_JTU = 0.975; β_JNE = 0.010) with a significant distance–decay in time (Mantel r = 0.172, p = 0.001; LM slope = 0.00128 per day ≈ 0.0128 per 10 days, R² = 0.069). For fungi, multi-site Jaccard β-diversity was extremely high (β_JAC = 0.970), dominated by turnover (β_JTU = 0.943; β_JNE = 0.027). Spatial distance–decay was negligible and non-significant (Mantel r = 0.087, p = 0.188; LM slope = 0.0050 per km, p = 0.336, R² = 0.008). Temporally, fungal β-diversity was highest of all (β_JAC = 0.995; β_JTU = 0.988; β_JNE = 0.0076) with a small but significant distance–decay in time (Mantel r = 0.046, p = 0.017; LM slope = 0.00110 per day ≈ 0.011 per 10 days, R² = 0.025).

Diversity patterns varied among groups: bacterial and plant diversity increased over time, whereas fungal diversity remained stable. Specifically, bacterial Shannon diversity increased significantly in common carder bee guts (t = 50, p = 0.01), with communities dominated by Gammaproteobacteria (genera ~~*XX*~~ [*Gilliamella*]{.underline} and [*Snodgrassella*]{.underline}), followed by Actinobacteria, Firmicutes, and Cyanobacteria (Fig. 1A). Notably, 9.13% of bacterial sequences could not be assigned to known taxa, suggesting the presence of novel or poorly characterized lineages (Fig. S2). Plant diversity also increased significantly (t = 6.26, p \< 0.001), while fungal communities showed little temporal change and were consistently dominated by Cladosporiales, Pleosporales, Eurotiales, Helotiales, and Saccharomycetales (Fig. 1B).
Furthermore, we found a positive relationship between the Shannon diversity of gut bacteria in common carder bees and the within-site component of floral species richness, indicating that gut microbial diversity increased at times when floral diversity at a given site was higher than its long-term mean. In contrast, the between-site component of floral richness and sampling period were not significant predictors, showing that differences in average floral diversity among sites and overall temporal trends did not explain additional variation in gut bacterial diversity once within-site fluctuations were accounted for. Sequencing depth did not alter these results, and the effect of within-site floral richness remained significant in models that accounted for temporal autocorrelation (AR(1)). These findings suggest that short-term, local increases in floral resource diversity are the main driver of gut microbial diversity in common carder bees, rather than broader spatial or long-term temporal differences (Fig. 3). In the case of fungi, we did not find any relationship between floral diversity either within or between-sites in the diversity of taxonomic groups (XXX). In the case of plants, we likewise found no association between plant species richness—either within (z= -0.93, p=0.35) or across sites (z=0.42, p=0.68)—and the richness of plant-derived gut contents; however, we detected a significant effect of period with plant species richness in gut contents generally increasing through time (z=3.91, p\<0.001, Fig. S8). In turn, gut plant richness had no detectable effect on bacterial ASV richness (z=-0.37, p=0.72, Fig. S9), but showed a significant positive relationship with fungal richness (z=3.31, p\<0.001, Fig. S10).

To explore potential associations among plants, bacteria, and fungi, we constructed a cross-domain correlation network. Because network inference is sensitive to sparse features, we first reduced each dataset to the most abundant taxa (100 plant taxa, 114 bacterial ASVs, and 104 fungal ASVs). The resulting network (Fig. 4, Fig. S11A-C) contained 12,795 associations (median Pearson correlation coefficient = –0.13, mean = 0.18), of which 27.8% were positive. By domain pair, we detected 4,653 plant–bacteria links (23.98% positive, 76.02% negative), 4,589 plant–fungi links (30.74% positive, 69.26% negative), and 3,553 bacteria–fungi links (28.91% positive, 71.09% negative). Across the cross-domain network, hub species included plants in the genera Lotus and Linaria, Bacteria in genera Gilliamella, Snodgrassella, and Enterobacterales, and Fungi included in Pseudogymnoascus, Antherospora, and Phialemonium taxonomic groups (Table S03).

In the comparison between early and late periods,, several microbial metabolic pathways showed increased relative abundance in the late period, suggesting functional shifts in the community. Notably, we observed an enrichment in multiple fermentation and menaquinone (vitamin K2) biosynthesis pathways, including the superpathway of fermentation (Chlamydomonas reinhardtii) and several superpathways of menaquinol biosynthesis (e.g., MK-6 to MK-13 variants, Fig. 5). These pathways are typically associated with anaerobic or facultative anaerobic metabolism, indicating a possible adaptation to lower oxygen availability or increased fermentation activity in the microbial community. Additionally, the increased abundance of the enterobacterial common antigen biosynthesis pathway points to a higher relative abundance of Enterobacteriaceae, a group often linked to environmental stress or host-associated niches. Enrichment in pyridoxal 5’-phosphate biosynthesis and salvage and L-aspartate/L-asparagine biosynthesis pathways suggests a rise in amino acid and cofactor biosynthesis, potentially reflecting metabolic specialization or nutrient demand. The presence of the chitin degradation II pathway may indicate greater abundance of chitinolytic bacteria, possibly in response to increased fungal biomass. Overall, these functional shifts point to a microbial community in the late period that is more metabolically versatile and adapted to anaerobic conditions, with increased capacity for nutrient processing and host/environment interaction.

## Discussion

Our analysis of the microbial composition of wild common carder bee guts shows that gut bacterial communities are dynamic across the flowering season and are shaped by both temporal variation and plant community characteristics. Similar to other previous research looking at the composition of gut microbiomes in other pollinator species, we find large differences throughout the flowering season and across the different sites sampled [@haag2022; @wang2019; @young2023].

Specifically, while common carder bees collected early in the season exhibit more depauperate patterns, featuring 27 to 68 ASVs, those captured late in the season host a more diverse microbial community (15 to 31). Our results indicate that it is within-site fluctuations in floral richness—rather than among-site differences or broad seasonal trends— that determine the observed increases in gut bacterial diversity. This pattern is consistent with two non-exclusive mechanisms: (i) a broader dietary niche exposes bees to a richer pool of floral microbes and metabolites (nectar/pollen chemistry) that can promote colonization by additional gut symbionts [@bosmans2018; @ptaszynska2021]; and (ii) greater plant diversity elevates microbial dispersal among flowers, pollinators and nests, increasing encounter rates with taxa that can persist in the gut [@lignon2024]. These findings highlight the dynamic nature of bumblebee gut microbiomes and underscore the importance of seasonal context when assessing microbial diversity and its ecological implications [@parmentier2016; @weinhold2024]. Further, this seasonal shift suggests a progressive accumulation of microbial diversity, potentially driven by increased foraging activity, exposure to diverse floral resources, and interactions with other pollinators as the season advances. The absence of a parallel signal for fungi suggests that the mycobiome is more environmentally filtered and site-specific. Our results indicate that within-site fluctuations in floral richness—rather than among-site differences or broad seasonal trends—track increases in gut bacterial diversity. 
When focusing on gut contents rather than site-level floral diversity, however, we detected the opposite pattern: bacterial richness was unrelated to gut plant diversity, whereas fungal richness increased significantly with the number of plant species consumed. These contrasting patterns suggest that bacterial communities are more strongly shaped by the broader floral diversity in the landscape, while fungal communities respond more directly to the diversity of plants actually consumed by bees. This highlights that different microbial groups may integrate plant resources at different ecological scales, reflecting distinct modes of acquisition and filtering within the host.

Common carder bee guts were colonized by a diverse array of bacteria, with Gammaproteobacteria being the most abundant, followed by Actinobacteria, Firmicutes, and Cyanobacteria. Notably, we observed a considerable proportion of unidentified bacterial taxa (9.13%,similar to [@bosmans2018a], reflecting both the complexity of microbial communities within pollinators [@koch2013; @hammer2021] and potential gaps in current taxonomic databases [@yuan2025]. This underlines the importance of continued exploration and sequencing efforts to fully characterize the microbial diversity associated with bumblebees.

In terms of temporal shifts, our results indicate the persistence of a core group of bacterial taxa consistently present across all time points and sampling sites. This stable core microbiome, which includes Lactobacillus, Gilliamella, Snodgrassella, and Bombiscardovia, likely plays fundamental roles in host health, such as enhancing nutrient absorption, protecting against pathogens, and modulating immune responses [@cariveau2014; @engel2016; @koch2011a; @mockler2018; @palmer-young2018]. The stability of this core community across both spatial and temporal scales suggests a functionally-conserved core involved in nutrient processing, pH regulation and pathogen exclusion [@engel2012a; @hammer2021] that may be evolutionarily conserved [@koch2013] . In contrast, late-season gains of Pseudomonas, Bacillus, Commensalibacter and Apibacter likely reflect transient, diet-linked colonizers acquired on flowers or nest material [@graystock2015; @bosmans2018a]. Such seasonally acquired associates may extend metabolic breadth or provide context-dependent benefits (e.g., carbohydrate utilization, detoxification, [@bonilla-rosso2018; @engel2012a], but their stability probably depends on continual environmental input [@kogan2023]. This late-season diversification could be driven by changes in floral availability and composition, shifts in foraging behavior, or increased environmental exposure as colonies expand and worker bees travel farther distances [@pope2018].

Our analysis revealed distinct temporal clustering of bacterial communities, with clear compositional differences between early and late time points. This temporal structuring suggests that the microbiome is not static but dynamically responds to seasonal changes in the environment. The early-season microbiome may be shaped by overwintering survival [@bosmans2018b] and the limited floral diversity available during initial foraging [@wood2022; @becher2024; @timberlake2024], while the late-season community reflects cumulative interactions with a broader spectrum of plant species [@keller2021; @vannette2020]. This shift may also be influenced by colony development stages [@su2021; @li2020], where increased colony size and worker numbers intensify environmental interactions.

In turn, fungal community composition varies mainly by site but shows limited temporal turnover, which implies strong local environmental filters (microclimate, surrounding vegetation, land use, nesting substrates) that shape which fungal spores or yeasts bees ingest and retain [@ptaszynska2021; @haag2022].

Across both bacterial and fungal domains, communities are highly heterogeneous and differences are overwhelmingly driven by turnover rather than nestedness, indicating species replacement (not simple richness loss) as the primary mechanism. This pattern is not inconsistent with the seasonal stability of a core microbiome: while dominant taxa remain relatively constant, much of the heterogeneity arises from the replacement of less abundant, transient taxa across space and time. Spatial separation explains little of the observed β-diversity—especially for fungi—suggesting substantial dispersal or strong local filtering that decouples composition from geographic distance at the sampled scales. By contrast, both bacteria and fungi show consistent, modest increases in turnover through time within sites, implicating temporal dynamics (e.g., environmental fluctuations, host/plant phenology, resource pulses) as more influential than pure space.

The cross-domain network was dominated by negative associations. Such patterns can be compatible with ecological processes like niche partitioning or antagonism among co-occurring taxa [@brochet2021], but given the correlational nature of the analysis, they should be interpreted cautiously and only as hypotheses of association [@kajihara2024]. Hubs highlighted biologically meaningful actors: plants in the genera Lotus and Linaria—common forage across sites—sat at the center of many cross-domain links; bacterial hubs included Gilliamella, Snodgrassella, and Enterobacterales-affiliated ASVs, matching their prominence in bumblebee guts; and fungal hubs featured Pseudogymnoascus, Antherospora, and Phialemonium. Together, these hubs likely represent key connector taxa that bridge diet-derived resources and microbial community dynamics. Because they have both high degree and strong associations, they may disproportionately influence how energy (e.g., from plant-derived sugars or pollen proteins) and information (e.g., metabolic signals, cross-feeding interactions) propagate through the gut microbiome. Identifying these central nodes highlights prime candidates for experimental validation, for instance by testing how diet manipulations or targeted colonization affect overall network structure and host–microbe interactions.

Our results suggest a functional shift in the gut microbiome between early and late periods, with the late-season community showing increased metabolic versatility—particularly in pathways related to fermentation, anaerobic metabolism, and nutrient processing—possibly reflecting adaptation to changing environmental conditions or host diet [@jones2017a; @ptaszynska2021; @bosmans2018].

Together, these findings underscore the dynamic nature of bumblebee gut microbiomes and the role of seasonal changes in shaping microbial community structure. They also highlight the importance of considering both temporal and spatial variation when assessing microbiome stability and its implications for host health and ecosystem functioning. Future studies with larger sample sizes and finer temporal resolution would be valuable for disentangling the drivers of microbiome shifts and their potential consequences for pollinator health and resilience. From a practical point of view, our results indicate that maintaining high, local floral diversity through time—rather than relying on site-to-site variation or brief mass-flowering events—may help support a richer, potentially more resilient gut microbiome in common carder bees. Given the site-sensitivity of the fungal assemblage, conservation actions should also consider site-level context (microclimate, floral composition, nesting habitat) when managing pollinator habitats.

**Acknowledgements**
Ainhoa Magrach acknowledges funding from the Ministry of Science, Innovation and Universities Grant (PID2021-127900NB-I00), the Basque Government through PIBA projects (2024RTE00060004), the European Union (ERC, GorBEEa 101086771), an Ikerbasque Research Professorship and the Spanish Ministry of Science and Innovation and the European Social Fund through the Ramón y Cajal Program (RYC2021-032351-I). Views and opinions expressed are, however, those of the author(s) only and do not necessarily reflect those of the European Union or the European Research Council. Neither the European Union nor the granting authority can be held responsible for them. This research is supported by María de Maeztu Excellence Unit 2023-2027 Ref. CEX2021-001201-M, funded by MCIN/AEI /10.13039/501100011033.


**Figure legends**

**Figure 1**. Genus-level composition of B. pascuorum gut microbiota across sampling periods during the flowering season in Gorbeia Natural Park. Relative abundance plots show the dominant A) bacterial or B) fungal genera detected through 16S rRNA and ITS1 metabarcoding respectively. Each bar represents the mean community composition of individuals sampled during each period, highlighting temporal shifts in microbial communities.

**Figure 2**. (A) Non-metric multidimensional scaling (NMDS) ordination of bacterial ASV composition across sampling periods. Points represent samples colored by period, with polygons indicating the convex hulls encompassing each group. The ordination illustrates shifts in microbial community composition through time. (B) Pairwise PERMANOVA results (Bray–Curtis dissimilarity) between sampling periods, with color intensity indicating the strength of group differences (–log10 FDR-adjusted p-values). Higher values (yellow) indicate stronger statistical support for differences between gut microbial communities, whereas lower values (blue/purple) indicate weak or non-significant differences. Together, the panels highlight temporal structuring of gut microbial communities in bumblebees.

**Figure 3**. Scatterplot showing the relationship between plant species diversity per site and bacterial ASV Shannon diversity across different time periods. Points are colored by sampling period to illustrate temporal variation in the association between plant and microbial diversity.

**Figure 4**. Tripartite cross-domain co-occurrence network (Plants \| Bacteria \| Fungi). Nodes are the top features retained per domain after read-depth filtering (≥300 reads) and CLR transformation; node size scales with degree. Plants are triangles, bacteria circles, and Fungi squares. Edges connect only across domains and indicate Pearson co-occurrence. Green lines indicate positive co-occurrence, red lines indicate negative co-occurrence; edge width is proportional to the Pearson coefficient of correlation The three domains are arranged in vertical columns and nodes are ordered by within-domain degree.

**Figure 5**. Differentially abundant microbial pathways between early and late sampling periods. Bar plots show the relative abundance of the top predicted metabolic pathways (y-axis) identified from functional profiling of microbial communities. Red bars correspond to the early period and blue bars to the late period; error bars indicate standard error of the mean. The log2 fold change in abundance between periods is shown to the right, and adjusted p-values (FDR-corrected) are listed in the final column. Pathways with lower adjusted p-values indicate stronger evidence for differential representation between early- and late-season communities.

**Figures**

```{r, include=TRUE, echo = FALSE, message=FALSE, error=FALSE, warning=FALSE, results='hide'}
# 
# #script 04_stats_ainhoa.R
# load("./results/RData/violin_plots_SHANNON_ASV_DIv.RData")
# 
# p1_sh


```

```{r, include=TRUE, echo = FALSE, message=FALSE, error=FALSE, warning=FALSE, results='hide'}

#script 04_stats_ainhoa.R
#load("./results/RData/combined_nmds.RData")

#combined


```

**Figure 3.**

```{r, include=TRUE, echo = FALSE, message=FALSE, error=FALSE, warning=FALSE, results='hide'}

#script 04_stats_ainhoa.R
#load("./results/RData/flower_richness_temporal_autocorrelation.RData")

# p4_sh2


```

**Figure 4.**

## References

##Supplementary material

```{r, include=TRUE, echo = FALSE, message=FALSE, error=FALSE, warning=FALSE, results='hide'}

#script 04_stats_ainhoa.R
#load("./results/RData/sampling_complete.RData")

# p


```

**Figure S1.**

```{r, include=TRUE, echo = FALSE, message=FALSE, error=FALSE, warning=FALSE, results='hide'}

#script 04_stats_ainhoa.R
#load("./results/RData/sampling_complete.RData")

# p

```

**Figure S1. Rarefaction and extrapolation curves showing ASV diversity across samples.** Sample-size-based rarefaction (solid lines) and extrapolation (dashed lines).

```{r, include=TRUE, echo = FALSE, message=FALSE, error=FALSE, warning=FALSE, results='hide'}
#script 04_stats_ainhoa.R
#load("./results/RData/barchart_order.RData")

# plot_order
```

**FIgure S2.**

```{r, include=TRUE, echo = FALSE, message=FALSE, error=FALSE, warning=FALSE, results='hide'}
#script 04_stats_ainhoa.R
#load("./results/nmds_site.RData")

# p2_site

```

**FIgure S3.**
