---
title: "The Changing Microbiome: How Seasonal Variation Shapes Bumblebee Gut Communities"
author: "Luis Javier Chueca^1^*, Jon Poza^1^, Manpreet Dhami^2,3^, Marion Donald^2^, Brais Hermosilla^1^, Jennifer Rose^1^, Xabier Salgado^1^, Ainhoa Magrach^1,4^"
corresponding-author: "Luis Javier Chueca (luisjavier.chueca@bc3research.org)"
format: docx
editor: visual
bibliography: references.bib
---

\small

```         
^1^Basque Centre for Climate Change-BC3, Edif. Sede 1, 1°, Parque Científico UPV-EHU, Barrio Sarriena s/n, 48940 Leioa, Spain

^2^Biocontrol & Molecular Ecology, Manaaki Whenua Landcare Research, Lincoln, New Zealand\

^3^School of Biological Sciences, Waipapa Taumata Rau, University of Auckland, Auckland, New Zealand\

^4^IKERBASQUE, Basque Foundation for Science, María Díaz de Haro 3,48013 Bilbao, Spain
```

## Abstract

Pollinators are essential for both natural and agricultural systems, supporting the diversity of plant communities and the productivity of crops around the globe. However, different global change drivers increasingly threaten these pollinator populations, leading to declines in population numbers and cascading effects on ecosystems and human food security. One critical but understudied aspect of pollinator health is their microbiome. The microbiome, a community of microorganisms living within and on pollinators, plays a significant role in digestion, immune function, and disease resistance, amongst others. However, environmental changes and pesticide exposure can disrupt these microbial communities, compromising pollinator health and survival. Most research on pollinator microbiomes has focused on managed species like honeybees, leaving significant gaps in our understanding of wild pollinators. This study investigates the microbiome composition of the common carder bee (*Bombus pascuorum*) across different floral environments throughout the flowering season. By analyzing the gut microbiomes of these bees, we aim to uncover how environmental factors shape their microbial communities, providing crucial insights into their resilience and adaptability in the face of global change. This knowledge is essential for guiding conservation efforts to protect pollinator populations.

## Introduction

Pollinators play a critical role in the functioning of both natural and agricultural systems worldwide, facilitating the reproduction of \>85% of flowering plants [@ollerton2011] and contributing to the production of \>75% of global food crops [@klein2006]. However, pollinators are increasingly at risk due to multiple global change drivers, something that has been recently defined as "*death by a thousand cuts*" [@wagner2021]. Climate change, with its impacts on temperature and precipitation patterns, disrupts pollinator behaviors and/or morphologies [@miller-struttmann2015] leading to plant-pollinator mismatches [@gérard2020]. Habitat loss and fragmentation lead to a reduction in pollinator populations [@winfree2009], while the widespread use of pesticides introduce additional stresses [@guzman2024]. These factors, compounded by the spread of diseases [@manley2015] and the increasing presence of invasive species [@vanbergen2017], threaten the survival of pollinator populations, leading to cascading effects on ecosystems and human food systems [@devkota2024; @murphy2022].

A relatively unexplored yet crucial aspect of pollinator populations is their microbiome [@shah2024], the community of microorganisms living within and on them. The microbiome can significantly impact the health, physiology [@Engel2013], development, and survival [@ferguson2018] of organisms, influencing digestion, immune function, and disease resistance. In insect pollinators, the gut microbiome plays a vital role in nutrient absorption [@zheng2017; @zheng2019], detoxification of harmful substances [@engel2012], and protection against pathogens [@steele2021; @koch2011]. Disruptions to the microbiome, whether through environmental changes, pesticide exposure, or other stressors, can compromise these essential functions and lead to increased vulnerability and mortality [@maes2016]. Understanding the composition and functionality of microbiomes is therefore critical, not only for gaining insights into the biology and ecology of pollinator species but also for developing strategies to mitigate the adverse effects of global change drivers.

Most research on pollinator microbiomes has focused on managed species, like honeybees [@engel2012] and certain bumblebee species [@weinhold2024], providing valuable insights into their gut microbial communities and how these microorganisms influence their health and productivity. However, our understanding of the microbiomes of other pollinator species, and particularly of wild species, remains limited (although see [@li2023] for a focus on other species). We know even less about how these microbiomes change over time or how different environmental factors shape them. As global change continues to alter ecosystems, understanding the temporal and spatial dynamics of pollinator microbiomes could provide valuable insights into how these species might cope with environmental stresses, maintain ecosystem functions, and contribute to biodiversity conservation.

Research in honeybees and some bumblebees has shown the existence of a core microbiome in these species, defined as a common group of microbes that are found across individuals of a given species (although the existence of this core microbiome is still somewhat controversial [@risely2020; @neu2021]. This core microbiome consists of around five genera in the case of honeybees [@carlini2024] and is mostly vertically-transmitted within the hive [@martinson2010]. In addition to this core microbiome, pollinators also acquire microbes from their environment, particularly through their diet [@ludvigsen2015], which can include nectar, pollen, and other floral resources. These acquired microorganisms can introduce new functions and adaptations but also make pollinators susceptible to environmental changes and stressors, which raises the need to understand the relationship between landscape composition (e.g., floral diversity) and microbiome composition.

Here, we focus on understanding the microbiome composition and dynamics of a wild pollinator species, the common carder bee (*Bombus pascuorum*), throughout the flowering season and across different sites that vary in floral diversity. By analyzing the gut microbiomes of these bees, we aim to uncover how the diversity and abundance of flower species in their environment influence the microbial communities they harbor. This study sheds light on the temporal changes in microbiome composition within *B. pascuorum* and reveals how different ecological contexts shape these vital microbial communities. Our findings contribute to a deeper understanding of the ecological interactions between wild pollinators and their environments, providing insights that are essential for conservation efforts and the maintenance of healthy pollinator populations.

## Methods

Our study was conducted within Gorbeia Natural Park, a protected area located in the Basque Country in N Spain. The region is shaped by the Atlantic Ocean's influence, resulting in mild winters and summers, alongside high annual rainfall, averaging around 1300 mm. During the study period (April to July), the average temperatures range from 10°C to 17.5°C [@Pizarro2021]. The area's vegetation includes beech forests mixed with extensive conifer plantations. Additionally, there is a notable presence of shrubland and meadows, particularly in the rockier, steeper regions of the mountain [@Albizu2002COMPARACINDL].

Within the park, we selected 16 different sampling sites, all of them located at similar altitudes \~800 m a.s.l, which we sampled every two weeks depending on weather conditions, avoiding rainy, windy or foggy days. At every sampling date we aimed to collect at least 5 *B. pascuorum* individuals per day, although this was not always possible. All collected individuals were labeled and stored independently in 50 ml vials at -20ºC until processed in the laboratory. For each bumblebee specimen we measured the inter-tegular distance as a measure of body size [@stemet2024]. We then extracted bumblebee guts using sterile forceps which we stored in 96% alcohol until DNA extraction. Bacterial, plant and fungi DNA was extracted following Macherey-Nagel NucleoSpin® 96 Soil protocol (2021) and were  amplified in duplicate using the DFD forward and ASDFAS reverse primers.

To avoid site and period vias all samples were randomized using a randomizer program before the DNA extraction, creating a new individual code for each sample.

*DNA Extraction and PCR Amplification*

Each gut sample was placed in a separate 2 mL vial with ceramic beads, and the ethanol was allowed to evaporate. Once dry, 700 µL of Lysis Buffer (SL1) were added to each vial, and the samples were mechanically homogenized (1750 rpm for 1 min with 15 s pauses, repeated 5 times). Contaminants were removed via precipitation, followed by centrifugation at 11,000 × g for 2 minutes to eliminate foam. Next, 150 µL of Buffer SL3 were added to each sample, vortexed for 5 seconds, and incubated on ice at 0-4ºC for 5 minutes, followed by centrifugation at 11,000 × g for 1 minute. The lysate was filtered using a square-well block and the NucleoSpin® Inhibitor Removal Plate. 550 µL of the clear supernatant was loaded into each well, and the plates were centrifuged at 2,250 × g for 13 minutes. The NucleoSpin® Inhibitor Removal Plate was discarded, and 250 µL of Buffer SB was added to the flow-through, mixed by pipetting. To bind the DNA, 725 µL of sample was transferred to a NucleoSpin® Soil Binding Plate and centrifuged at 2,250 × g for 13 minutes. The flow-through was discarded. DNA was washed in four steps: (1) 500 µL of Buffer SB, centrifuged at 2,250 × g for 13 minutes; (2) 550 µL of Buffer SW1, centrifuged at 2,250 × g for 5 minutes; (3) 700 µL of Buffer SW2, centrifuged at 2,250 × g for 5 minutes; (4) 700 µL of Buffer SW2, centrifuged at 2,250 × g for 13 minutes. After washing, the plate was incubated at 37ºC for 20 minutes to dry the silica membrane. DNA was eluted by adding 100 µL of Buffer SE to each well, incubating for 1 minute, and centrifuging at 2,250 × g for 5 minutes. A total of 132 samples were obtained, with each sample labeled accordingly. Nanodrop tests were performed on random samples to confirm successful DNA extraction.

*PCR Amplification*

DNA amplification was assessed using two-step PCR. Step 1 PCRs were performed for bacterial, plant, and fungal genetic material, with identical reagent volumes for all samples, differing only in the primers used (Table xxx). PCRs were run in a 96-well plate, with two plates for each group: one containing 96 samples and another with 40 samples, including negative controls. Step 1 PCR was repeated twice per sample, after which 7.5 µL from each replicate were pooled into a new plate, with 15 µL per well. PCR amplification success was confirmed via 2% agarose gel electrophoresis between Step 1 and Step 2 PCRs to verify amplicon size.

*Statistical analyses*

To evaluate the completeness of our surveys, we used the iNEXT package [@iNEXT], which provides robust estimates of species diversity and sampling coverage based on sample-size- and coverage-based rarefaction and extrapolation. We compared the observed sample coverage with the estimated asymptotic coverage to quantify the fraction of the community captured by our sampling efforts.

We analyzed community composition using Non-metric Multidimensional Scaling (NMDS) with the Bray-Curtis dissimilarity index, implemented in the vegan package in R [@vegan]. Species abundance data were square-root transformed to reduce the influence of dominant species. We then performed a Permutational Multivariate Analysis of Variance (PERMANOVA) with 999 permutations to assess significance. The time period and the site in which the specimens were collected were specified as the main factor.

To assess the relationship between floral species diversity and the gut microbiome of wild bumblebees, we modeled three response variables separately: bacterial species richness (observed ASVs), Shannon diversity, and Simpson diversity. For each of these diversity metrics, we fitted a linear model with floral diversity as a predictor, while controlling for potential variation due to sampling period and site by including period and site as covariates.

*Functional Prediction of Microbial Communities*
To infer the functional potential of the bacterial communities, we used PICRUSt2 (Phylogenetic Investigation of Communities by Reconstruction of Unobserved States, version X.X) to predict metagenomic profiles from 16S rRNA gene sequence data. Amplicon sequence variants (ASVs) were aligned to a reference phylogeny using the place_seqsand hsp.py steps, and functional profiles were inferred based on KEGG Orthology (KO) and MetaCyc pathway databases. Predicted pathway abundances were normalized and compared across time periods. Based on observed differences in bacterial diversity and composition over time, we grouped sampling periods into early (March-May) and late (May-July) categories. Statistical comparisons of pathway relative abundances between these two groups were performed using differential abundance analysis, with significance assessed via permutation tests and adjusted p-values. This approach allowed us to assess seasonal shifts in the functional potential of gut microbial communities associated with changes in environmental conditions and floral resources.


## Results

Throughout the flowering season, we collected a total of 132 bumblebee specimens across 16 sites, with an average of XX ± SD individuals per site and sampling period. Sampling completeness was generally high, averaging 90% across sites and periods, although it ranged widely from 21% to full completeness (Fig. S1), indicating robust coverage of the bacterial communities in our samples.
Our results demonstrate a significant increase in bacterial diversity within bumblebee guts over time, as measured by Shannon diversity indices (Fig. 1; t = 50, p = 0.01). This suggests a dynamic shift in the gut microbiome throughout the season. The bacterial communities were dominated by Gammaproteobacteria, particularly genera XX and XX, followed by representatives of Actinobacteria, Firmicutes, and Cyanobacteria. Notably, a relatively large fraction (X%) of the bacterial sequences remained unidentified, pointing to the presence of potentially novel or poorly characterized taxa (Fig. S2).

Examining temporal changes in community composition revealed a stable core microbiome consistently present across all time periods (Fig. 3) and sampling sites (Fig. SX). This core includes key ASVs from well-known bumblebee-associated genera such as Lactobacillus, Gilliamella, Snodgrassella, and Bombiscardovia, likely reflecting essential symbionts fundamental to host health and function. Interestingly, as the flowering season progressed, this core microbiome was supplemented by additional bacterial taxa, including species from genera such as Pseudomonas, Bacillus, Commensalibacter, and Apibacter, indicating a seasonal enrichment or colonization of the gut by a broader suite of bacteria.

Multivariate analyses of bacterial community composition through time showed clear temporal structuring, with distinct clustering of samples by sampling period. The most pronounced compositional differences occurred between early and late-season samples, supported by a significant PERMANOVA test (Fig. 4; F = 2.3111, p = 0.002), underscoring a notable shift in gut microbiota as the season advances.

Furthermore, we found a positive relationship between bacterial diversity in bumblebee guts and local plant species richness, indicating that sites with higher floral diversity support richer gut microbial communities. This association may partly reflect the overall increase in floral species diversity observed through time (Fig. 5), suggesting that the diversity of available floral resources influences the gut microbiome diversity of pollinators.

Our preliminary analyses of fungal communities similarly indicate dynamic changes over the season, 

In the comparison between early and late periods, several microbial metabolic pathways showed increased relative abundance in the late period, suggesting functional shifts in the community. Notably, we observed an enrichment in multiple fermentation and menaquinone (vitamin K2) biosynthesis pathways, including the superpathway of fermentation (Chlamydomonas reinhardtii) and several superpathways of menaquinol biosynthesis (e.g., MK-6 to MK-13 variants, Fig. 7). These pathways are typically associated with anaerobic or facultative anaerobic metabolism, indicating a possible adaptation to lower oxygen availability or increased fermentation activity in the microbial community. Additionally, the increased abundance of the enterobacterial common antigen biosynthesis pathway points to a higher relative abundance of Enterobacteriaceae, a group often linked to environmental stress or host-associated niches. Enrichment in pyridoxal 5’-phosphate biosynthesis and salvage and L-aspartate/L-asparagine biosynthesis pathways suggests a rise in amino acid and cofactor biosynthesis, potentially reflecting metabolic specialization or nutrient demand. The presence of the chitin degradation II pathway may indicate greater abundance of chitinolytic bacteria, possibly in response to increased fungal biomass. Overall, these functional shifts point to a microbial community in the late period that is more metabolically versatile and adapted to anaerobic conditions, with increased capacity for nutrient processing and host/environment interaction.

## Discussion

Our analysis of the microbial composition of wild bumblebee guts shows that gut bacterial communities are dynamic across the flowering season and are shaped by both temporal variation and plant community characteristics.

Similar to other previous research looking at the composition of gut microbiomes in other bumblebee species, we find large differences throughout the flowering season (REF). Specifically, while bumblebees collected early in the season exhibit more depauperate patterns, featuring XX to XX ASVs, those captured later on in the season host a much more diverse microbial community (XXX to XXX). This seasonal shift suggests a progressive accumulation of microbial diversity, potentially driven by increased foraging activity, exposure to diverse floral resources, and interactions with other pollinators as the season advances. These findings highlight the dynamic nature of bumblebee gut microbiomes and underscore the importance of seasonal context when assessing microbial diversity and its ecological implications.

Bumblebee guts were colonized by a diverse array of bacteria, with Gammaproteobacteria being the most abundant, followed by Actinobacteria, Firmicutes, and Cyanobacteria. Notably, we observed a considerable proportion of unidentified bacterial taxa (**X%**), reflecting both the complexity of microbial communities within pollinators and potential gaps in current taxonomic databases. This underlines the importance of continued exploration and sequencing efforts to fully characterize the microbial diversity associated with bumblebees.

In terms of temporal shifts, our results indicate the persistence of a core group of bacterial taxa consistently present across all time points and sampling sites. This stable core microbiome, which includes Lactobacillus, Gilliamella, Snodgrassella, and Bombiscardovia, likely plays fundamental roles in host health, such as enhancing nutrient absorption, protecting against pathogens, and modulating immune responses (REF). The stability of this core community across both spatial and temporal scales suggests a strong host-microbe association that may be evolutionarily conserved.

Interestingly, as the season progressed, the core microbiome was complemented by additional bacterial taxa, particularly species within the genera Pseudomonas, Bacillus, Commensalibacter, and Apibacter. This late-season diversification could be driven by changes in floral availability and composition, shifts in foraging behavior, or increased environmental exposure as colonies expand and worker bees travel farther distances (REF). These supplementary taxa may represent transient members of the microbiome that are acquired from specific floral resources or environmental sources, potentially enhancing metabolic capabilities or stress tolerance during the late foraging season.

Our analysis revealed distinct temporal clustering of bacterial communities, with clear compositional differences between early and late time points. This temporal structuring suggests that the microbiome is not static but dynamically responds to seasonal changes in the environment. The early-season microbiome may be shaped by overwintering survival and the limited floral diversity available during initial foraging, while the late-season community reflects cumulative interactions with a broader spectrum of plant species. This shift may also be influenced by colony development stages, where increased colony size and worker numbers intensify environmental interactions.

Our results suggest a functional shift in the gut microbiome between early and late periods, with the late-season community showing increased metabolic versatility—particularly in pathways related to fermentation, anaerobic metabolism, and nutrient processing—possibly reflecting adaptation to changing environmental conditions or host diet.

Together, these findings underscore the dynamic nature of bumblebee gut microbiomes and the role of seasonal changes in shaping microbial community structure. They also highlight the importance of considering both temporal and spatial variation when assessing microbiome stability and its implications for host health and ecosystem functioning. Future studies with larger sample sizes and finer temporal resolution would be valuable for disentangling the drivers of microbiome shifts and their potential consequences for pollinator health and resilience.

Figure legends


**Figure 1.** *Temporal patterns of bacterial ASV diversity across periods.*  Violin plots show the distribution of observed ASV diversity values for each period, with embedded boxplots indicating medians and interquartile ranges. The dashed grey line represents the linear trend over time, highlighting directional changes in microbial diversity across periods. Diversity estimates were standardized using interpolation and extrapolation techniques to control for differences in sampling effort.

**FIgure 2**. Relative abundance of major bacterial taxa in bumblebee gut samples across six sampling periods (p01 to p06). Each point represents the percentage of 16S rRNA gene copies recovered for a given taxon in an individual sample. Boxplots summarize the distribution of relative abundances per taxon, highlighting dominant groups such as Gammaproteobacteria and Unclassified bacteria. Temporal variation in community composition is indicated by different colors and shapes corresponding to sampling periods.

**FIgure 3**.

**Figure 4.** Non-metric multidimensional scaling (NMDS) ordination of bacterial ASV composition across sampling periods. Points represent samples colored by period, with polygons indicating the convex hulls encompassing each group. The plot visualizes differences in microbial community composition over time.

**Figure 5.** Scatterplot showing the relationship between plant species diversity and bacterial ASV diversity across different time periods. Points are colored by sampling period to illustrate temporal variation in the association between plant and microbial diversity.

**Figures**


```{r, include=TRUE, echo = FALSE, message=FALSE, error=FALSE, warning=FALSE, results='hide'}

#script explore data
load("./results/RData/violin_plots_SHANNON_ASV_DIv.RData")

p1_sh


```
**Figure 1**


```{r, include=TRUE, echo = FALSE, message=FALSE, error=FALSE, warning=FALSE, results='hide'}

#script explore data
load("./results/RData/nmds.RData")

p2_hull


```
**Figure 5.**

```{r, include=TRUE, echo = FALSE, message=FALSE, error=FALSE, warning=FALSE, results='hide'}

#script explore data
load("./results/RData/scatter_plant_ASV_Div.RData")

p4_sh


```
**Figure 6.**



## References


##Supplementary material


```{r, include=TRUE, echo = FALSE, message=FALSE, error=FALSE, warning=FALSE, results='hide'}

#script explore data
load("./results/RData/sampling_complete.RData")

p


```

**Figure S1.**

**Figure S1. Rarefaction and extrapolation curves showing ASV diversity across samples.** Sample-size-based rarefaction (solid lines) and extrapolation (dashed lines).

